---
title: "03-measure-static-allometry"
author: "Ariel Marcy"
date: "2019-01-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Static Allometry
Changing size is one of the most common ways that organisms can change their shape. Modifications to growth during development often have a profound impact on adult shape. The tests in this script detect how much size appears to drive shape change in our sample.

This script generates Figure 1 (and data for Table 1a). 

### Load packages, functions, and data
```{r message = FALSE}
library(stringr)
library(data.table)
library(geomorph)
library(colorspace)  # colors
source("../Functions/utilities.R")  # loads custom functions
load(file = "../Data/Processed/02-main-data.rda")
```

### Set up for analyses
Since *Mastacomys* is technically within the genus *Pseudomys* (Smissen & Rowe 2018), we'll make an "Effective Genus" column where this genus is classified under *Pseudomys*.
```{r}
info$EGenus <- info$Genus  # make Effective Genus column
info$EGenus[which(info$Genus == "Mas")] <- "Pse"  # Mas becomes Pseudomys
```

Colors for subsequent figures are set up according to column "EGenus". Colors roughtly correspond to phylogenetics such that similar color hues indicate closer evolutionary relationships.
```{r}
# Make gradients for each clade
Uro <- sequential_hcl(5, "Greens 3")  # Uromyini clade
Con <- sequential_hcl(5, "Purples 3")  # Conilurus clade
Pse <- sequential_hcl(5, "Blues 3")  # Pseudomys clade
Hyd <- sequential_hcl(5, "Reds 3")  # Hydromini clade
Pog <- sequential_hcl(5, "Heat 2")  # Pogonomys
Inv <- sequential_hcl(5, "Light Grays")  # Mus and Rattus

# Colors in order of phylogeny
genera.phylo <- c("Mus", "Rat", "Pog", "Xer", "Hyd", "Not", "Pse", "Zyz", "Leg", "Uro", "Mel", "Lep", "Mes", "Con")

col.phylo <- c(Inv[3], Inv[1], Pog[3], Hyd[1], Hyd[2], Pse[1], Pse[2], Pse[4], Pse[3], Uro[1], Uro[2], Con[3], Con[2], Con[1])

names(col.phylo) <- genera.phylo

# Colors in order of alphabetical genus names
col.gen <- col.phylo[match(sort(unique(info$EGenus)), names(col.phylo))]
sp.col.gen <- col.phylo[match(as.factor(info$EGenus), names(col.phylo))]  # color vector for all data
```

## 1) Measure static allometry: Homogeneity of Slopes Test by species
Isolate only those species with at least 8 specimens.
```{r}
spec.counts <- as.data.frame(table(info$Taxa))
to.remove <- spec.counts[which(spec.counts$Freq <= 7), ]
species.rm <- as.vector(to.remove$Var1)

# For loop to take out these specimens
dataset.rm <- NULL  # initiate blank data frame
for(n in 1:dim(to.remove)[1]) {
        dataset.rm <- c(dataset.rm, which(info$Taxa == species.rm[n]))
}

# Remove species from:
info.HOS <- info[c(-dataset.rm), ]  # Metadata
shape.HOS <- shape[, , c(-sort(dataset.rm))]  # Shape dataset
```

Run allometry tests as well as a homogeneity of slopes test. These tests take about 5 min to run. 
```{r}
# Run static allometry test
spp.gdf <- geomorph.data.frame(shape = shape.HOS, spp = info.HOS$Taxa, size = info.HOS$Csize)
spp.results <- procD.allometry(shape ~ log(size), f2 = ~spp, iter = 499, data = spp.gdf)

# Examine results
spp.results$aov.table
spp.results$HOS.test
summary(spp.results)

# Find pair-wise p-values for slopes for each taxa
size <- info.HOS$Csize
spp <- info.HOS$Taxa
pwHOS.spp <- advanced.procD.lm(f1 = shape.HOS ~log(size) + spp, 
                                     f2 = ~log(size) * spp, 
                                     groups = ~spp,  
                                     slope = ~log(size),
                                     iter = 499,
                                     angle.type = "deg")

# Export p-value matrix to Results folder
write.csv(pwHOS.spp$P.slopes.dist, "../Data/Results/TableS4_HOS_test.csv")
```

## 2) Visualize Static Allometry: Predicted Value
In _geomorph_, static allometry can be visualized with log(centroid size) plotted versus the projected regression score of shape ~ size, a common way of quantifying "size-less" shape.
```{r}
# Run static allometry test with all specimens
spp.gdf <- geomorph.data.frame(shape = shape, Species = info$Taxa, size = info$Csize)
spp.results <- procD.allometry(shape ~ log(size), f2 = ~Species, iter = 499, data = spp.gdf)

# Set colors for genera and different points for species within each genus
pch.unique.spp <- PointOutDiffSpp(info)  # points by unique species in a genus for legend
pch.unique.spp[which(sort(unique(info$Taxa)) == "M_fus")] <- 2
pch.spp <- PlotByGroup(info, "Taxa", pch.unique.spp)  # pch coded for each individual

# Make color vector for legend
unique.taxa <- as.data.frame(unique(info[, c(6,10)]))
unique.taxa <- na.omit(unique.taxa[order(unique.taxa), ])
col.unique.spp <- PlotByGroup(unique.taxa, "EGenus", col.gen)

# Put color and point legends in phylogenetic order
phylo.names <- c("M_mus", "R_rat", "R_nor", "R_lut", "R_leu", "R_tun", "R_vil", "R_sor", "R_fus", "P_mol", "X_myo", "H_chr", "N_cer", "N_fus", "N_mit", "N_ale", "P_pat", "P_gra", "P_ora", "P_sho", "P_des", "P_apo", "P_aus", "P_hig", "P_her", "P_nov", "P_del", "M_fus", "Z_arg", "L_for", "U_cau", "M_cer", "M_cap", "M_bur", "L_con", "M_gou", "C_pen")
phylo.order <- na.omit(match(phylo.names, unique.taxa$Taxa))
names(col.unique.spp) <- NULL
col.unique.spp <- col.unique.spp[phylo.order]
pch.unique.spp <- pch.unique.spp[phylo.order]

# Find the mean point for each static allometry
mean.size <- aggregate(spp.results$size, list(Taxa = info$Taxa), mean)
mean.shape <- aggregate(spp.results$Reg.proj, list(Taxa = info$Taxa), mean)
```

### Plot/export static and evolutionary allometry multipanel Figure 1
Here we consolidate the information above into two static allometry plots, one with all the data points and one with just the regression lines for each species to compare with the evolutionary allometry of the entire group. 
```{r}
setEPS()  # sets up plot export
postscript("../Data/Results/Figure2_Static_Evo_Allo.eps")
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))

# Plot of all data points with evolutionary allometry
par(mar = c(1, 4, 1.5, 0))
plot(x = log(spp.results$size),
     y = spp.results$Reg.proj,
     xlim = c(4.7, 6.4),
     col = sp.col.gen, 
     pch = pch.spp,
     bg = sp.col.gen,
     main = "Static and evolutionary allometry",
     xlab = NA,
     ylab = "Shape (Regression Score)")
text(min(log(spp.results$size)), max(spp.results$Reg.proj), "a")
abline(lm(mean.shape$V1 ~ log(mean.size$x)), col = "dark grey", lwd = 2)  # plots evolutionary allometry
text(5.7, -0.04, paste("Rsq of log(size) = ", round(spp.results$aov.table$Rsq[1], 2)), cex = 1.1, col = "dark grey")

# Species Legend
legend(6.045, 0.06, legend = phylo.names, col = col.unique.spp, border = NULL, pch = pch.unique.spp, pt.bg = col.unique.spp, cex = 0.97, pt.cex = 1, ncol = 2, bg = "white")

# Plot of static allometry with evolutionary allometry
par(mar=c(5, 4, 1.5, 0))
plot(x = log(spp.results$size),
     y = spp.results$pred.val,
     xlim = c(4.7, 6.4),
     col = sp.col.gen, 
     pch = pch.spp,
     bg = sp.col.gen,
     xlab = "Log(centroid size)",
     ylab = "Shape (Predicted Value)")
text(min(log(spp.results$size)), max(spp.results$pred.val), "b")
text(5.6, -0.08, paste("Rsq of log(size):species = ", round(spp.results$aov.table$Rsq[3], 2)), cex = 1.1, col = "dark grey")
dev.off()
```

### Save intermediate data
```{r}
save(col.gen, sp.col.gen, col.phylo, phylo.names, spp.results, file = "../Data/Processed/03-color-data.rda")
```