---
title: "04-test-evolutionary-allometry"
author: "Ariel Marcy"
date: "2019-01-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Evolutionary allometry with Procrustes ANOVA
Evolutionary studies must take into account the phylogenetic tree because related species do not behave the same as independent samples. 

This script generates Table 1.

### Load packages, functions, and data
```{r message = FALSE}
library(stringr)
library(geomorph)
library(ape)  # needed to work with phylogenies
library(geiger)  # needed to work with phylogenies
library(phylotate)  # needed to work with phylogenies
library(data.table)
library(vegan)
source("../Functions/utilities.R")  # loads custom functions
load(file = "../Data/Processed/02-main-data.rda")
load(file = "../Data/Processed/03-color-data.rda")
```

## Set up shape array for evolutionary allometric and phylogentic analyses
First, attach long form names for genus and species to the metadata.
```{r}
# Get names from trait information gathered from Native Mice and Rats by Dr.s Bill Breed and Fred Ford (2007). 
traits <- read.csv("../Data/Processed/in_ex_traits.csv", header = TRUE)
traits$Taxa <- paste(str_sub(traits$Genus, 1, 1), str_sub(traits$Species, 1, 3), sep = "_")  # make matching Taxa column
info.traits <- merge(info, traits, by = "Taxa", sort = F)
info.traits <- info.traits[order(info.traits$Order), ]  # preserve shape specimen order

# Rename Genus & Species columns so they have their old, short column name, long form Genus and Species columns keep Genus.y and Species.y
names(info.traits)[names(info.traits) == 'Genus.x'] <- 'Genus'
names(info.traits)[names(info.traits) == 'Species.x'] <- 'Species'

info.phylo <- info.traits[, -c(5:7)]  # remove unnecessary columns
```

Second, rename shape dataset dimnames to only "Genus_species". This is so we can re-order the shape data to match the tree later on. (We can still reference the CatNum in the metadata.)
```{r}
shape.phylo <- shape  # initiate shape optimized for phylogenetics
dimnames(shape.phylo)[[3]] <- as.list(paste(info.phylo$Genus.y, info.phylo$Species.y, sep = "_"))
```

## Calculate mean shape and centroid size for each species
This step has to be done before plotting evolutionary allometry and working with the phylogeny.

The mean shape function, `mshape()` outputs a 2D array. Therefore, the `for` loop below stores mean shapes in a 2D array format recognized by `geomorph`: (n x [p x k]), where n is number of specimens, p is number of landmark points, and k is number of dimensions. Finally, we convert to a 3D array with format (p x k x n) since a 3D array is required by most `geomorph` functions. 
```{r}
info.phylo$FullName <- paste(info.phylo$Genus.y, info.phylo$Species.y, sep = "_")  # give names to metadata in the same format as the tree, "Genus_species"
full.names <- unique(info.phylo$FullName)  # list of species

# For loop to subset shape data by species and calculate mean shape
mean.shapes.2D <- NULL  # initiate blank data frame
for (i in 1:length(full.names)) {  
        is.taxa <- which(info.phylo$FullName == full.names[i])
        shape.for.mean.2D <- shape.phylo[, , is.taxa]
        mean.shapes.2D <- rbind(mean.shapes.2D, mshape(shape.for.mean.2D))  # output cols must = 3 for 3D
}
mean.shapes <- arrayspecs(mean.shapes.2D, 325, 3)  # convert to 3D array with 325 landmarks and 3 for 3D landmarks in this study
```

Create metadata and put mean shape and metadata in alphabetical order.
```{r}
# Alphabetical order for mean shape
dimnames(mean.shapes)[[3]] <- full.names  # Genus_species format
mean.shapes <- mean.shapes[, , sort(dimnames(mean.shapes)[[3]])]  # ABC

# Alphabetical order for mean shape metadata
info.means <- info.phylo[!duplicated(info.phylo$Taxa), -c(4:6)]  # make metadata; remove outdated columns: CatNum, Csize, Order
info.means <- info.means[order(info.means$FullName), ]  # ABC order
```

### Calculate mean centroid size
We need to update the centoid size column for info.means to contain the *mean* centroid size by taxa. Centroid size for all 317 specimens is stored in the info.phylo metadata table.
```{r}
# Calculate mean centroid size and store in info.means
cs.means <- aggregate(info.phylo[, 6], list(Taxa = info.phylo$FullName), mean)  # aggregate also alphabetizes it
info.means$MeanCsize <- cs.means[, 2]  # store in vector
```

## Prepare the tree data for phylogenetic analyses
The commands in `geomorph` require the tree and datasets to have exactly the same set of species.

### Load in phylogenetic tree
The most recent tree for Australian rodents is from Smissen & Rowe 2018. The tree is a fossil calibrated ultrametric tree from BEAST2 for Hydromini, which includes all the Australo-Papuan old endemics in Murinae. This tree adds 6 more species from Australia to the Smissen & Rowe phylogeny. 
```{r}
raw.tree <- read_annotated("../Data/Processed/Marcy-BEAST01.con.tre")
```

### Extract only Genus and Species names from the tree
The tree file has uneven amounts of information for each entry, some entries have catalog numbers and others do not. Therefore we needed a `for` loop to grab the first two elements from a list, which always corresponded to Genus and Species.

We reassign only Genus and Species names (in the form "Genus_species") to the tree so that it is easier to compare to the shape data. 
```{r}
no.cat.tree <- raw.tree  # initiate new, modified tree with no CatNums
label.list <- strsplit(raw.tree$tip.label, "_")  # separate info into list

# For loop finds Genus & Species and makes tip name "Genus_species"
for (i in 1:length(label.list)) {
        label <- unlist(label.list[i])
        no.cat.tree$tip.label[i] <- paste(label[1], label[2], sep = "_")
}
```

### Re-label node
Since the 2018 Smissen & Rowe tree focused primarily on New Guinea species, it does not have all of the Australian species in our shape dataset. However, it does include sister species with the same relative branch length to other genera in the study. Therefore, we decided to re-label one of the nodes to the sister Australian species. 

**Relabeled node:**
_Pogonomys loriae_ (NG) to _Pogonomys mollipilosus_ (Aus)

```{r}
relabel.tree <- no.cat.tree
P.lor <- which(str_detect(no.cat.tree$tip.label, "Pogonomys_loriae"))
relabel.tree$tip.label[P.lor] <- paste("Pogonomys_mollipilosis")
```

### Prune the tree and shape datasets to match each other
Now that the names in the Smissen & Rowe tree can be compared to our info.phylo metadata, we can prune both datasets to have the same species.

First, prune the tree of species not in the shape data
```{r}
concord <- name.check(relabel.tree, relabel.tree$tip.label, dimnames(mean.shapes)[[3]])  # check concordance
aus.tree <- drop.tip(relabel.tree, concord$tree_not_data)  # prune tree to Aus sample
```

Next, subset and re-order the metadata and shape datasets to match the order of species found in the tree. 
```{r}
# Subsetting
to.remove <- name.check(aus.tree, aus.tree$tip.label, dimnames(mean.shapes)[[3]])
info.means.tree <- info.means[!is.element(info.means$FullName, to.remove$data_not_tree), ]  # metadata
mean.shapes.tree <- mean.shapes[, , !is.element(info.means$FullName, to.remove$data_not_tree)]  # shape data

# Re-order to match tree tip order
tree.order <- match(aus.tree$tip.label, info.means.tree$FullName)
info.means.tree <- info.means.tree[tree.order, ]  # metadata
mean.shapes.tree <- mean.shapes.tree[, , tree.order]  # shape data
```

## Visualize tree for Figure 1
Phylogeny with colors coordinated to major clades (see script 03)
```{r}
# Colors by species, ordered in same way as tree
grp <- as.factor(str_sub(aus.tree$tip.label, 1, 3))
grp[which(grp == "Mas")] <- "Pse"
names(col.gen) <- sort(unique(grp))
col.tree <- col.gen[match(grp, names(col.gen))]

# Make Figure 1
setEPS()
postscript("../Data/Results/Figure1_Phylogeny.eps")
plot.phylo(aus.tree, edge.color = "gray", tip.color = col.gen, no.margin = T)
dev.off()
```

## Evolutionary Allometry, reported in Table 1
Here we run Procrustes ANOVAs to compare the relative impact of genus, size, and their interaction on shape.
```{r}
# Procrustes ANOVA with interaction
evo.allo.gen.gdf <- geomorph.data.frame(Shape = mean.shapes.tree, size = info.means.tree$MeanCsize, Genus = info.means.tree$Genus)

evo.allo.gen <- procD.pgls(Shape ~log(size) * Genus, aus.tree, iter = 499, data = evo.allo.gen.gdf)
evo.allo.gen
```

### Export Tables 1a,b as CSV files
```{r}
# Round all values to 3 decimals and remove NAs from the tables
table1a <- as.data.frame(round(spp.results$aov.table, 3))
table1a[is.na(table1a)] <- ""

table1b <- as.data.frame(round(evo.allo.gen$aov.table, 3))
table1b[is.na(table1b)] <- ""

# Export
write.table(table1a, "../Data/Results/Table1a_species_ANOVA.csv", sep = ",", col.names = NA)
write.table(table1b, "../Data/Results/Table1b_genus_ANOVA.csv", sep = ",", col.names = NA)
```

### Save intermediate data
```{r}
save(relabel.tree, mean.shapes.tree, info.means.tree, aus.tree, file = "../Data/Processed/04-phylo-data.rda")
```