---
title: "05-measure-evolutionary-rates"
author: "Ariel Marcy"
date: "2/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Evolutionary rates
Evolutionary studies must take into account the phylogenetic tree because related species do not behave the same as independent samples. 

### Load packages, functions, and data
```{r message = FALSE}
library(dispRity)
library(geomorph)
library(geiger)
source("../Functions/utilities.R")  # loads custom functions
load(file = "../Data/Processed/02-main-data.rda")
load(file = "../Data/Processed/04-phylo-data.rda")
```

## Compare evolutionary rates
Given the New World monkey papers (Chervaud & Marriog 2005, 2009), we would expect specialists like carnivores to have the slowest rates of evolution and rodents that evolve along the allometric line to have the greatest rates of evolution. 

First, we need to make sure the data meets the brownian motion assumption required by the Kmult method, based on Blomberg's K (Blomberg et al. 2003), 
```{r} 
# Plot disparity through time
mean.shapes.tree.2D <- two.d.array(mean.shapes.tree)
aus.dtt <- dtt(aus.tree, mean.shapes.tree.2D, nsim = 500, plot = TRUE)
```
Data mostly fits the Brownian Motion model with one possible exception around relative time ~0.65.

**_Notomys_ versus other species**
```{r}
# Set up grouping factor
gp.notomys <- rep(0, dim(info.means.tree)[1])  # set "other" as 0
gp.notomys[which(info.means.tree$Genus == "Not")] <- 1  # set Notomys
names(gp.notomys) <- info.means.tree$FullName  # required to run analysis

# Run analysis
not.test <- compare.evol.rates(mean.shapes.tree, aus.tree, gp.notomys, iter = 500)
summary(not.test)
```

**Carnivores vs non-carnivores**
```{r}
is.carnivore <- as.numeric(info.means.tree$Feeding == "Carnivorous")
names(is.carnivore) <- info.means.tree$FullName
carn.test <- compare.evol.rates(mean.shapes.tree, aus.tree, is.carnivore, iter = 500)
```
This test showed no significant difference in evolutionary rates.

**Herbivores vs other species**
```{r}
# Only specialized herbivores P. oralis and M. fuscus
P.ora <- which(info.means.tree$Taxa == "P_ora")
M.fus <- which(info.means.tree$Taxa == "M_fus")
is.sp.herb <- c(P.ora, M.fus)
herb.grps <- rep(2, dim(info.means.tree)[1])
herb.grps[is.sp.herb] <- 1

names(herb.grps) <- info.means.tree$FullName
herb.test <- compare.evol.rates(mean.shapes.tree, aus.tree, herb.grps, iter = 500)
```

**Large frugivores vs others**
```{r}
# Only large frugivores U. caudimaculatus and M. gouldii
U.cau <- which(info.means.tree$Taxa == "U_cau")
M.gou <- which(info.means.tree$Taxa == "M_gou")
is.frug <- c(U.cau, M.gou)
frug.grps <- rep(2, dim(info.means.tree)[1])
frug.grps[is.frug] <- 1

names(frug.grps) <- info.means.tree$FullName
frug.test <- compare.evol.rates(mean.shapes.tree, aus.tree, frug.grps, iter = 500)
```

### Adujust p-values for multiple comparisons
```{r}
p.adjust(c(not.test, carn.test$P.value, herb.test$P.value, frug.test$P.value), method = "bonferroni")
```