---
title: "05-plot-heatmaps"
author: "Ariel Marcy"
date: "3/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Plot PC1 Min/Max
To make the landmark variation heatmaps, we use Dr Thomas Guillerme's new package, `landvR` to see how specific landmarks vary along PC1. 

```{r message = FALSE}
library(png)  # needed to work with screenshots
library(geomorph)
# Load libraries (and install if necessary) for landvR package
if(!require(devtools)) install.packages("devtools")
if(!require(dispRity)) install.packages("dispRity")
if(!require(landvR)) install_github("TGuillerme/landvR")
source("../Functions/utilities.R")  # loads custom functions
load(file = "../Data/Processed/02-main-data.rda")
```

### Set up
The below was adapted from a [vignette written by Dr Guillerme](https://cdn.rawgit.com/TGuillerme/landvR/8a6a6bd5/inst/vignettes/Landmark_partition_test.html).
```{r}
# Find mean configuration - i.e. mean shape in dataset
consensus <- (select.procrustes(shape, selector = mean))[[1]]

# Measure spherical coordinates differences from the mean for each specimen for each landmark
diff.from.mean <- coordinates.difference(coordinates = shape, reference = consensus, type = "spherical")

# Ordinate the data
twoD.shape <- two.d.array(shape)
ordination <- stats::prcomp(twoD.shape)

# Force the symmetric component of shape into class "gpagen" (required for variation.range)
gpagen.shape <- list()
gpagen.shape$coords <- shape
gpagen.shape$consensus <- consensus
class(gpagen.shape) <- "gpagen"

# Measure extremes of variation from mean on PC1
PC1.var <- variation.range(gpagen.shape, return.ID = FALSE, axis = 1, ordination = ordination, type = "spherical")

# Wrap specimens on the tangent space
wrap.PCA <- plotTangentSpace(shape, verbose = FALSE)

# Select extreme specimens (the ones that make the warp-meshes)
hypothetical.1 <- wrap.PCA$pc.shapes[[1]]
hypothetical.2 <- wrap.PCA$pc.shapes[[2]]
```

## Plot the range of variation along PC1 using a heat color scheme
`3d` commands allow precise control of 3D visualization to get dorsal and lateral views

ERROR: rgl.snapshot() can't save the png files to the Data/Results folder so the screenshots end up in the Analysis folder
```{r}
# Dorsal view
open3d(windowRect = c(0, 0, 600, 600))
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = heat.colors, col.val = PC1.var[, "radius"], labels = F)
view3d(phi = 0)  # flip cranium
rgl.snapshot("PC1heatmap_dor.png")  # save screenshot for multipanel plot

# Lateral view
open3d(windowRect = c(0, 0, 600, 600))
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = heat.colors, col.val = PC1.var[, "radius"], labels = F)
view3d(phi = 90)
rgl.snapshot("PC1heatmap_lat.png")
```

### Prep images for multi-panel figure
This step assumes you have ImageMagik installed on your computer
```{r}
system("mogrify PC1heatmap_*.png -trim *.png")  # removes whitespace (requires ImageMagik)
dor <- readPNG("PC1heatmap_dor.png")
lat <- readPNG("PC1heatmap_lat.png")
```

## Make multipanel Figure 3
ERROR: memory issues when trying to add legend within setEPS. Works when lines 86-103 are run through the console, plot is "zoomed" and then screenshot from that window
```{r}
setEPS()  # sets up plot export
postscript("../Data/Results/Figure3_Heatmaps.eps")
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))  # 2 rows, 2 columns
par(mar=c(0, 0, 0, 0))

# Plot dorsal view
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T) 
rasterImage(dor, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "a", cex = 2)

# Plot lateral view with same dimensions as above
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T)
rasterImage(lat, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "b", cex = 2)

# Make gradient legend
legend_image <- as.raster(matrix(heat.colors(20), ncol = 1))
rasterImage(legend_image, 440, 220, 465, 160)
text(393, 213, "High variation", col = "dark grey", cex = 1.4)
text(420, 168, "Low", col = "dark grey", cex = 1.4)
dev.off()
```