---
title: "05-plot-heatmaps"
author: "Ariel Marcy"
date: "2/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Plot allometry heatmaps
Here we plot the shape differences across PC1 and PC2 to detect what shape changes occur. PC1 corresponded to predicted allometric shape. 

We use Dr Thomas Guillerme's new package, `landvR` to see how specific landmarks vary along the PC axes. Much of the below was adapted from [this vignette written by Dr Guillerme](https://cdn.rawgit.com/TGuillerme/landvR/8a6a6bd5/inst/vignettes/Landmark_partition_test.html).

### Load packages, functions, and data from previous steps
```{r message = FALSE}
# Load libraries (and install if necessary) for landvR package
if(!require(devtools)) install.packages("devtools")
if(!require(dispRity)) install.packages("dispRity")
if(!require(landvR)) install_github("TGuillerme/landvR")
library(data.table)
library(vegan)
library(geomorph)
source("../Functions/utilities.R")  # custom functions
load(file = "../Data/Processed/02-main-data.rda")
```

### Plot PC1 Min/Max
```{r}
# Find mean configuration - i.e. mean shape in dataset
consensus <- (select.procrustes(shape, selector = mean))[[1?]]

# Measure spherical coordinates differences from the mean for each specimen for each landmark
diff.from.mean <- coordinates.difference(coordinates = shape, reference = consensus, type = "spherical")

# Ordinate the data
twoD.shape <- two.d.array(shape)
ordination <- stats::prcomp(twoD.shape)

# Force the symmetric component of shape into class "gpagen" (required for variation.range)
gpagen.shape <- list()
gpagen.shape$coords <- shape
gpagen.shape$consensus <- consensus
class(gpagen.shape) <- "gpagen"

# Measure extremes of variation from mean on PC1
PC1.var <- variation.range(gpagen.shape, return.ID = FALSE, axis = 1, ordination = ordination, type = "spherical")

# Wrap specimens on the tangent space
wrap.PCA <- plotTangentSpace(shape, verbose = FALSE)

# Select extreme specimens (the ones that make the warp-meshes)
hypothetical.1 <- wrap.PCA$pc.shapes[[1]]
hypothetical.2 <- wrap.PCA$pc.shapes[[2]]

# Plot the range of variation along PC1 using a heat color scheme
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = heat.colors, col.val = PC1.var[, "radius"], labels = F)
```

### Plot PC2 Min/Max (not working for this version of landvR)
```{r}
# Measure extremes of variation from mean on PC2
PC2.var <- variation.range(gpagen.shape, return.ID = FALSE, axis = 2, ordination = ordination, type = "spherical")

# Wrap specimens on the tangent space; flip axes to get correct warp grids
wrap.PCA.2 <- plotTangentSpace(shape, verbose = FALSE, axis1 = 1, axis2 = 2)

# Select extreme specimens (the ones that make the warp-meshes)
hypothetical.1.2 <- wrap.PCA.2$pc.shapes[[1]]
hypothetical.2.2 <- wrap.PCA.2$pc.shapes[[2]]

# Plot the range of variation along PC2 using a heat color scheme
PC1.var.plot <- procrustes.var.plot(hypothetical.1.2, hypothetical.2.2, col = heat.colors, col.val = PC2.var[, "radius"], labels = F)
```