---
title: "05-plot-evolutionary-allometry"
author: "Ariel Marcy"
date: "2/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Phylogenetic analyses
Here we plot the Principal Components Analysis as a phylomorphospace to visualize morphological variation in context with the phylogeny. PC1 is highly correlated allometric shape. 

We also plot evolutionary allometry to visualize species which diverge.

This script generates Figure 2. 

### Load packages, functions, and data from previous steps
```{r message = FALSE}
library(stringr)
library(plotrix)  # needed to draw ellipses
library(ape)  # needed to work with phylogenies
library(geiger)  # needed to work with phylogenies
library(data.table)
library(geomorph)
source("../Functions/utilities.R")  # custom functions
load(file = "../Data/Processed/02-main-data.rda")
load(file = "../Data/Processed/03-color-data.rda")
load(file = "../Data/Processed/04-phylo-data.rda")
```

## Make multipanel Figure 2: mean shape PCA and evolutionary allometry
First, set up colors and points by genera and important taxa
```{r}
# Make an Effective Genus column to categorize Mastacomys phylogenetically within Pseudomys
info.means.tree$EGenus <- info.means.tree$Genus
info.means.tree$EGenus[which(info.means.tree$Genus == "Mas")] <- "Pse"

# Colors for genera and different points by species within each genus; used for PCA
col.means <- PlotByGroup(info.means.tree, "EGenus", col.gen[2:length(col.gen)])  # standard colors from script 03 minus genus Conilurus
pch.means <- PointOutDiffSpp(info.means.tree)  # pts by unique species in a genus
pch.means[which(info.means.tree$Taxa == "M_fus")] <- 2  # open triangle

# Points to call out only specialized herbivores with special characters; used for Evolutionary Allometry plot
pch.evo.allo <- rep(16, length(col.means))  # other taxa are circles
pch.evo.allo[which(info.means.tree$Taxa == "P_ora")] <- 11  # star
pch.evo.allo[which(info.means.tree$Taxa == "M_fus")] <- 2  # open triangle
```

### Run PCA
```{r}
pca.means <- plotTangentSpace(mean.shapes.tree)  # PCA of mean shapes

# Write x and y labels with proportion of variance for PC1 and PC2
PCs <- pca.means$pc.summary$importance
PC1.per <- round(PCs[2, 1] * 100, digits = 1)  # % with 1 decimal
PC1.lab <- paste("PC", 1, " (", PC1.per, "%)", sep = "")
PC2.per <- round(PCs[2, 2] * 100, digits = 1)
PC2.lab <- paste("PC", 2, " (", PC2.per, "%)", sep = "")
```

### Generate data for Evolutionary Allometry plot
```{r}
# Run allometry on all 34 species
evo.procD <- geomorph.data.frame(shape = mean.shapes.tree, size = info.means.tree$MeanCsize, genus = info.means.tree$Genus)
mean.allo <- procD.allometry(shape ~ log(size), data = evo.procD)
```

### Plot and export the multi-panel Figure 2
PCA on top and evolutionary allometry plot on the bottom

```{r}
aus.rates <- as.numeric(aus.rates)
greyPal <- colorRampPalette(c("white", "black"))
gray.branch <- greyPal(10)[cut(aus.rates, breaks = 5)]

plotGMPhyloMorphoSpace(aus.tree, mean.shapes.tree, tip.labels = F, node.labels = F, plot.param = list(t.bg = col.means, t.pch = pch.means, l.col = gray.branch))

```


```{r}
setEPS()  # sets up plot export
postscript("../Data/Results/Figure2_PCA_EvoAllo.eps")
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))  # 2 rows, 2 columns

# PCA plot
plot(x = -pca.means$pc.scores[, 1],
     y = pca.means$pc.scores[, 2],
     xlim = c(-0.08, 0.12),
     ylim = c(-0.08, 0.12),
     col = col.means,
     pch = pch.means,
     bg = col.means,
     main = "a) Principal Component Analysis",
     xlab = PC1.lab, 
     ylab = PC2.lab)
text(.105, -0.003, "Frugivores", col = "dark grey")
draw.ellipse(.116, .015, .006, .012, angle = -12, border = "dark grey")

text(.085, 0.041, "Herbivores", col = "dark grey")
draw.ellipse(.048, .035, .025, .013, angle = -19, border = "dark grey")

text(.045, -0.054, "Carnivores", col = "dark grey")
draw.ellipse(.044, -.073, .034, .007, angle = -15, border = "dark grey")

text(-.055, 0.035, "Notomys", col = "dark grey")
draw.ellipse(c(-.053, -0.07), c(.014, -0.013), c(.015, .004), c(.004, .007), angle = c(-95, 0), border = c("dark grey", "dark grey"))  # draws both

# Evolutionary Allometry plot
plot(x = log(info.means.tree$MeanCsize),
     y = mean.allo$Reg.proj,
     xlim = c(4.75, 6),
     col = col.means, 
     pch = pch.evo.allo,
     main = "b) Evolutionary Allometry",
     xlab = "Log centroid size", 
     ylab = "Shape (regression score)")
legend(5.81, 0.052, legend = names.phylo, col = col.phylo, pch = 16, cex = 1, pt.cex = 1, ncol = 2)

abline(lm(mean.allo$Reg.proj ~ log(info.means.tree$MeanCsize)), col = "dark grey", lwd = 1)  # plots evolutionary allometry line of best fit
text(4.81, 0.105, paste("r = ", cor.allo.short), col = "dark grey")

text(5.75, 0.11, "Frugivores", col = "dark grey")
draw.ellipse(5.92, .115, .08, .005, angle = 1, border = "dark grey")

text(5.35, 0.08, "Herbivores", col = "dark grey")
draw.ellipse(5.41, .047, .11, .02, angle = 5, border = "dark grey")

text(5.8, 0.066, "Carnivore", col = "dark grey")
text(5.03, 0.015, "Carnivore", col = "dark grey")

text(5.3, -0.065, "Hopping Notomys", col = "dark grey")
draw.ellipse(5.12, -0.055, .075, .012, angle = 8, border = "dark grey")

dev.off()  # finishes plot export
```