---
title: "06-plot-evolutionary-allometry"
author: "Ariel Marcy"
date: "2/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Phylogenetic analyses
Here we plot the Principal Components Analysis as a phylomorphospace to visualize morphological variation in context with the phylogeny. PC1 is highly correlated allometric shape. 

We also plot evolutionary allometry to visualize species which diverge.

This script generates Figure 3. 

### Load packages, functions, and data from previous steps
```{r message = FALSE}
library(stringr)
library(ape)  # needed to work with phylogenies
library(geiger)  # needed to work with phylogenies
library(phytools)  # needed to work with phylogenies
library(data.table)
library(geomorph)
library(plotrix)  # needed to draw ellipses

load(file = "../Data/Processed/02-main-data.rda")
load(file = "../Data/Processed/03-color-data.rda")
load(file = "../Data/Processed/04-phylo-data.rda")
source("../Functions/utilities.R")  # custom functions
source("../Functions/plotGMPhyloMorphoSpace_plotmod.R")  # modified geomorph function
source("../Functions/phytools.branch.colors.R")  # modified phytools function
```

## Make multipanel Figure 3: evolutionary allometry and mean shape PCA/phylomorphosapce
First, set up colors and points by genera and important taxa
```{r}
# Make an Effective Genus column to categorize Mastacomys phylogenetically within Pseudomys
info.means.tree$EGenus <- info.means.tree$Genus
info.means.tree$EGenus[which(info.means.tree$Genus == "Mas")] <- "Pse"

# Colors for genera and different points by species within each genus; used for PCA
col.means <- PlotByGroup(info.means.tree, "EGenus", col.gen)  # standard colors from script 03 minus genus Conilurus
pch.means <- PointOutDiffSpp(info.means.tree)  # pts by unique species in a genus
pch.means[which(info.means.tree$Taxa == "M_fus")] <- 2  # open triangle

# Points to call out only specialized folivores with special characters; used for Evolutionary Allometry plot
pch.evo.allo <- rep(16, length(col.means))  # other taxa are circles
pch.evo.allo[which(info.means.tree$Taxa == "P_ora")] <- 11  # star
pch.evo.allo[which(info.means.tree$Taxa == "M_fus")] <- 2  # open triangle
```

### Run PCA
```{r}
pca.means <- plotTangentSpace(mean.shapes.tree)  # PCA of mean shapes

# Write x and y labels with proportion of variance for PC1 and PC2
PCs <- pca.means$pc.summary$importance
PC1.per <- round(PCs[2, 1] * 100, digits = 1)  # % with 1 decimal
PC1.lab <- paste("PC", 1, " (", PC1.per, "%)", sep = "")
PC2.per <- round(PCs[2, 2] * 100, digits = 1)
PC2.lab <- paste("PC", 2, " (", PC2.per, "%)", sep = "")
```

### Generate data for Evolutionary Allometry plot
```{r}
# Run allometry on all 34 species
evo.procD <- geomorph.data.frame(shape = mean.shapes.tree, size = info.means.tree$MeanCsize, genus = info.means.tree$Genus)
mean.allo <- procD.allometry(shape ~ log(size), data = evo.procD)
```

### Isolate molecular rates used for phylogeny in paper and color-code branches by rates
```{r}
annotations <- aus.tree$node.comment

# Use function to strip rates from annotated NEXUS file
StripAnnotation <- function(input, annotation) {
        strsplit(strsplit(input, split = annotation)[[1]][2], split = ",")[[1]][1]
}
rates <- as.character(lapply(as.list(annotations), StripAnnotation, annotation = ",rate="))

aus.rates <- as.numeric(rates[which(!is.element(relabel.tree$tip.label, concord$tree_not_data))])
names(aus.rates) <- aus.tree$tip.label

# Plot molecular rate of evolution, light branch = slower; dark =faster
gray.branches <- phytools.branch.colors(aus.tree, aus.rates, gray)
gray.branches.10 <- darken(gray.branches, 0.1)  # make white branches visible
```

## Plot and export the multi-panel Figure 3
Evolutionary allometry plot on top, PCA in the middle, two heatmaps on the bottom

ERROR: memory issues when trying to add legend within setEPS. Works when lines 107-177 are run through the console, plot is "zoomed" from RStudio, and then screenshot from that window
```{r}
# Find Notomys' color
Not.col <- col.gen[which(names(col.gen) == "Not")]

# Read in heatmap PNGs created with script 05
dor <- readPNG("../Data/Processed/PC1heatmap_dor.png")
lat <- readPNG("../Data/Processed/PC1heatmap_lat.png")

setEPS()  # sets up plot export
postscript("../Data/Results/Figure3_EvoAllo_PCA_Heatmaps.eps")
mat <- matrix(c(1,1,2,2,3,4), 3, 2, byrow = TRUE)  # 3 rows, 2 columns
layout(mat, widths = c(1, 1, 1), heights = c(0.65, 0.75, 0.5))  

# 1) Evolutionary Allometry plot
par(mar = c(4, 4, 1, 1))  # sets the margins
plot(x = log(info.means.tree$MeanCsize),
     y = mean.allo$Reg.proj,
     xlim = c(4.75, 6),
     col = col.means, 
     pch = pch.evo.allo,
     xlab = "Log centroid size", 
     ylab = "Shape",
     bty = "n")
text(4.75, 0.1, "a", cex = 2)

legend(x = "bottomright", legend = genera.phylo, col = col.phylo, pch = 16, cex = 0.85, pt.cex = 0.85, ncol = 4, bty = "n")

abline(lm(mean.allo$Reg.proj ~ log(info.means.tree$MeanCsize)), col = "dark grey", lwd = 1)  # plots evolutionary allometry line of best fit

text(5.77, 0.11, "Frugivores", col = "dark grey")
draw.ellipse(5.92, .115, .08, .005, angle = 1, border = "dark grey")

text(5.35, 0.08, "Folivores", col = "dark grey")
draw.ellipse(5.41, .047, .11, .02, angle = 5, border = "dark grey")

text(5.8, 0.066, "Carnivore", col = "dark grey")
text(5.05, 0.015, "Carnivore", col = "dark grey")

text(5.3, -0.055, "Hopping Notomys", col = Not.col)
draw.ellipse(5.12, -0.055, .075, .012, angle = 8, border = "dark grey")

# 2) PCA
par(mar=c(4, 4, 1, 1))  # sets the margins
PCAlims = c(-0.08, 0.12)
plot(x = -pca.means$pc.scores[, 1],
     y = pca.means$pc.scores[, 2],
     xlim = PCAlims,
     ylim = PCAlims,
     col = col.means,
     pch = pch.means,
     bg = col.means,
     xlab = PC1.lab, 
     ylab = PC2.lab,
     bty = "n")
text(-0.08, 0.1, "b", cex = 2)
text(.105, -0.009, "Frugivores", col = "dark grey")
draw.ellipse(.116, .015, .006, .012, angle = -12, border = "dark grey")
text(.045, 0.065, "Folivores", col = "dark grey")
draw.ellipse(.048, .035, .025, .013, angle = -19, border = "dark grey")
text(.045, -0.053, "Carnivores", col = "dark grey")
draw.ellipse(.044, -.073, .034, .007, angle = -15, border = "dark grey")
text(-.067, 0.034, "Notomys", col = Not.col)
draw.ellipse(c(-.053, -0.07), c(.014, -0.013), c(.015, .004), c(.004, .007), angle = c(-95, 0), border = c("dark grey", "dark grey"))  # draws both

# 3) Plot dorsal view
par(mar = c(0, 0, 1, 0))
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T) 
rasterImage(dor, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "c", cex = 2)

# 4) Plot lateral view with same dimensions as above
par(mar = c(0, 0, 1, 0))
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T)
rasterImage(lat, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "d", cex = 2)

# Make gradient legend
legend_image <- as.raster(matrix(heat.colors(20), ncol = 1))
rasterImage(legend_image, 440, 220, 465, 160)
text(382, 213, "High variation", col = "dark grey", cex = 1)
text(414, 168, "Low", col = "dark grey", cex = 1)

dev.off()
```

### Plot phylomorphospace for supplementary materials
Uses modified `geomorph` function `plotGMPhyloMorphoSpace_plotmod` which can be found in the Functions folder of this repository.
```{r}
gray.branches.10 <- rep("dark gray", length(gray.branches.10)) # needed handle decision to no longer color the branches

setEPS()
postscript("../Data/Results/FigureS3_PhyloPCA.eps")

plotGMPhyloMorphoSpace_plotmod(aus.tree, mean.shapes.tree, xaxis = -1, tip.labels = F, node.labels = F, plot.param = list(t.bg = col.means, t.pch = pch.means, t.cex = 1.5, n.cex = .75, l.col = gray.branches.10, lwd = 2, main = "PCA Phylomorphospace", xlab = PC1.lab, ylab = PC2.lab, xlim = PCAlims, ylim = PCAlims))

dev.off()  # finishes plot export
```