---
title: "plot-allometry-no-log"
authors: "Ariel Marcy"
date: "2/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../allometry-rodents')
```

# Visualize allometry
This script re-generates Figure 2, the static and common allometry plots without taking the log of centroid size.

### Load packages, functions, and data from previous steps
```{r message = FALSE}
library(geomorph)
load(file = "../Data/Processed/03-main-data.rda")
load(file = "../Data/Processed/03-graphing-vectors.rda")
```

## Figure 2: Static and common allometric slopes with centroid size (no log)
In _geomorph_, static allometry can be visualized with centroid size plotted versus the projected regression score of shape ~ size, a common way of quantifying "size-less" shape. All static alloemtries can be compared to a common evolutionary allometry using the mean point for each static allometry:
```{r}
# Re-do allometry with no log (from script 04)
nolog.gdf <- geomorph.data.frame(coords = shape, size = info$Csize, spp = info$Taxa)
fit.nolog <- procD.lm(coords ~ size * spp, iter = 500, data = nolog.gdf, print.progress = FALSE)

# Re do regression score
nolog.allo.rs <- plotAllometry(fit.nolog, size = info$Csize, logsz = FALSE, method = "RegScore")  # Allometry data

# Re do finding means
nl.mean.size <- aggregate(nolog.allo.rs$size.var, list(Taxa = info$Taxa), mean)
nl.mean.shape <- aggregate(nolog.allo.rs$RegScore, list(Taxa = info$Taxa), mean)
```

### Plot/export static and evolutionary allometry multi-panel Figure 2
Here we consolidate the information above into two static allometry plots, one with all the data points and one with just the regression "pred-lines" for each species to compare with the evolutionary allometry of the entire group. 
```{r}
setEPS()  # sets up plot export
postscript("../Data/Results/Figure2_Static_Allo_nolog.eps")
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))

# Plot of all data points with evolutionary allometry
par(mar = c(1, 4, 1.5, 0))
plot(fit.nolog,
     type = "regression",
     reg.type = "RegScore",
     predictor = info$Csize,
     col = sp.col.gen, 
     pch = pch.spp,
     bg = sp.col.gen,
     main = "Static and evolutionary allometry (no log)",
     xlab = NA,
     ylab = "Shape (Regression Score)",
     bty = "n")
text(4.7, 0.048, "A", cex = 2)
abline(lm(nl.mean.shape$V1 ~ nl.mean.size$x), col = "dark grey", lwd = 2)  # plots evolutionary allometry
text(250, -0.04, paste("Rsq of size = ", round(fit.nolog$aov.table$Rsq[1], 2)), cex = 1.1, col = "black")

# Species Legend
legend(6.05, 0.06, legend = phylo.names, col = col.unique.spp, border = NULL, pch = pch.unique.spp, pt.bg = col.unique.spp, cex = 0.97, pt.cex = 1, ncol = 2, bg = "white")

# Plot of static allometry with evolutionary allometry
par(mar = c(5, 4, 1.5, 0))
plot(fit.nolog,
     type = "regression",
     reg.type = "PredLine",
     predictor = info$Csize,
     col = sp.col.gen, 
     pch = pch.spp,
     bg = sp.col.gen,
     xlab = "Centroid size",
     ylab = "Shape (Predicted Value)",
     bty = "n")
text(4.7, 0.11, "B", cex = 2)
text(250, -0.08, paste("Rsq of size:species = ", round(fit.nolog$aov.table$Rsq[3], 2)), cex = 1.1, col = "black")

dev.off()
```